---
title: "Data Expedition: Using R to investigate the epidemiology of parasites in geladas"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This lab introduces the basics of data visualization in R. In the exercises at the end, you will develop your own hypotheses and visually explore patterns of infection and mortality in a dataset on gelada monkeys and tapeworms. 

You should have already gone to the [R website](https://cran.r-project.org/) and installed the appropriate version of R for your operating system. (if you missed class on Monday, please do this step now!)

## Introducing R and data visualization

In this section, we will generate artificial data and produce some basic visualizations in R. You will later apply these techniques with real data in the exercises. Simple modifications to this code will be sufficient to complete the exercises at the end.   

To begin, open R on your computer. You will see a window that you can type into much like a calculator: this is the R command line. Try typing and `2 + 2` onto the command line and hit `enter`:

```{r}
2 + 2
```

Throughout this document, you can copy and paste the blocks of R code highlighted with `gray boxes` into your own R console to mimic what I'm doing. The white boxes that begin with a '##' are the output from R that you should see in your R console if you enter the code in the gray box (e.g., entering '2 + 2' into R produces the answer '4', as seen above in the white box). Keep in mind that some blocks of code only work if you have already entered earlier blocks of code. Lines beginning with a hashtag are 'comments' to help us keep track of what the code is doing, and are ignored by R.

A key feature of R is the ability to define and manipulate 'objects'. Objects can be individual numbers or character strings, vectors of numbers of character strings, "data frames" which are tables of data similar to an Excel spreadsheet, and much more. To define an object in R, we can use the '=' sign, which is also called an 'assignment operator' because it assigns an object to a name. For example, below I define several types of objects in R (note that the old 'x' is overwritten each time we use the assignment operator):

```{r}
# a number can be an object
x = 2
x

# a character string can be an object
x = "dog"
x

# a vector of character strings (enclosed in quotes) can be an object
x = c("dog", "cat", "bird")
x

# a vector of numbers can be an object
x = c(1, 2, 3, 4, 5)
```

Once an object is defined in your R session, you can use R functions to do things to that object. R has many built in functions for manipulating data and for performing mathematical and statistical operations. For example, here are some things we can do to the vector of numbers 'x' created above::

```{r}
# add them up
sum(x)

# add +1 to each value
x + 1

# log transform each value
log(x)

# raise each value to the power of 3
x^3

# find the maximum value in the vector
max(x)

# turn them into characters
as.character(x)

# randomly sample 10 values from x with replacement
sample(x, 10, replace=TRUE)
```

Now, let's use some special R functions to generate fake data that we will use to demonstrate some of R's visualization functions. Copy and paste the following block of code into R and hit `enter`. Don't worry about how this code works- you won't need to do anything like this for the exercise:  

```{r}
# generate random data with 20 rows
set.seed(0)
fakedata = data.frame(
  color = sample(c("blue", "green"), 20, replace=T),
  direction = sample(c("up", "down"), 20, replace=T),
  size = rnorm(20),
  speed = rnorm(20))
```

Now, let's look at the object `fakedata` we just created by entering its name on the R command line:

```{r}
fakedata
```

We have two *categorical variables* (color and direction), and two *continuous variables* (size and speed). There are several types of plots we can make with this data. First, use the `attach` function to 'attach' the data in your R session- this is just a convenient trick that allows us to refer to individual columns of the data by simply typing the column name (some warning messages may pop up on your R console, but it's nothing to worry about!):

```{r}
# attach data
attach(fakedata)

# test it out by typing the name of one of the columns in fakedata
color
```

We can visualize the distribution of *a single numerical variable* with a **histogram**. This is useful for answering questions like "are values skewed? what is the average? how scattered are the values?" For example, we can look at the histogram for "size" in our fake data, and see that it is centered around zero, it doesn't seem to be skewed, and it ranges from about -1.5 to 1.5:

```{r}
# histogram of 'size'
hist(size)
```


We can visualize the distribution of *a single categorical variable* with a barplot. For example, we can make a barplot for the 'color' variable that compares the frequency of green vs. blue. To do this, we first need a table that shows the number of data points falling in each category (green or blue), and then pass this table to the barplot function:

```{r}
# create table (just type 'tab1' into R to see what it looks like)
tab1 = table(color)

# make barplot
barplot(tab1)
```

We can visualize the distributions of *one numerical variable across different levels of a categorical variable* with a **boxplot**. This is useful if we want to compare the distribution of a variable across two or more categories, for example if we want to see whether the distribution of size is different for "blue" verses "green":

```{r}
# boxplot of 'size' verses 'color'
boxplot(size ~ color)
```

We can visualize the *relationship between a pair of numerical variables* with a **scatterplot**. Oftentimes, we expect variables to have a positive or negative association. Let's look at the relationship between "size" and "speed" in our data:

```{r}
# scatterplot of 'size' verses 'speed'
plot(size ~ speed)
```

Finally, we can look at the *number of items falling into a category across different levels of another category* using a ** side by side barplot**. For example, with our fake data, we could investigate whether the amount of 'green' is higher when the direction is 'up' compared to when the direction is 'down'. As with the barplot for a single categorical variable, we need to use the 'table' function to make a contingency table showing

```{r}
# make a table of color vs direction
tab2 = table(color, direction)

# look at the table
tab2
```

Now we can feed this table to the `barplot` function. In this case, the setting `beside=TRUE` tells R to put the bars side-by-side rather than stacked, which is the default. The code below also illustrates how to specify additional graphical parameters, separated by commas, within the plot function to specify aesthetic features, such as colors for the bars (`col=c("blue", "green"`) and labels to put below the bars (`names.arg=c("down", "up")`):

```{r}
# make barplot showing distribution of colors for up vs. down
barplot(tab2, beside=TRUE, col=c("blue", "green"), names.arg=c("down", "up"))
```

You can see that the barplot corresponds to the values in the table: the lowest value, 4, is for 'down, blue', and the highest value, 6, is for 'up, green'. 

Now that you have seen the basics, below is code for two plots that demonstrate a few more of the common graphical parameters you may want to use in order to customize the look of your plots, particularly customizing axis labels and adding a legend to the barplot:

```{r}
# pink histogram with customized axis and plot labels
hist(size, col="pink", xlab="Custom x-axis label", ylab="Custom y-axis label", main="Custom main title for plot")

# barplot
barplot(tab2, beside=TRUE, col=c("darkblue", "forestgreen"), names.arg=c("Down", "Up"), ylab="Count")

# add legend (you may have to tinker with the x and y coordinates for the legend when you make your own figures)
legend(x=0.5, y=7, legend=c("Blue", "Green"), fill=c("darkblue", "forestgreen"), bty="n", xpd=TRUE)
```

To learn more about the types of settings you can use for each plot function, type a question mark followed by the function name into the R console and the help page for that function will pop up:

```{r}
# pull up R help pages (don't worry if these are a bit confusing)
?plot
?boxplot
?barplot
```

To find more valid R colors, just Google "colors in R". 

## Exercise: exploring data on tapeworm parasitism in geladas

This exercise is divided into 3 parts. For each part, you must develop a hypothesis and prediction about the data, produce a visualization that is relevant to your hypothesis, and then discuss whether the visualization supports your hypothesis.

Before beginning the exercise, you must load the data into R. You can download the data directly into your R session by entering the following line of code:

```{r}
#data = source('')
```

Use the `head` function to look at the first few rows of data, including the column names:

```{r}
#head(data)
```

[describe data]

Finally, let's attach the data as we did for the fake data we made before. Now you are ready to start the exercise!

```{r}
#attach(data)
```

### Part 1: Geladas reach adulthood around age 5. Do cysts tend to appear before or after geladas reach adulthood?

1a. State a hypothesis for whether cysts tend to appear in younger or older adults (you can state a hypothesis for *either* cysts appearing in younger *or* older adults, but choose one):

1b. State the prediction that flows from your hypothesis.

1c. Present a data visualization that tests your hypothesis (e.g., histogram, scatterplot, boxplot, or barplot).

1d. Describe your visualization in 2-3 sentences. Does it support your hypothesis?

```{r setup, include=FALSE, echo=FALSE}
hist(cyst_age)
```

### Part 2: Do cysts appear earlier for males or females?

1a. State a hypothesis for whether cysts appear earlier in males or females (you can state a hypothesis for *either* cysts appearing earlier in males *or* females, but choose one):

1b. State the prediction that flows from your hypothesis.

1c. Present a data visualization that tests your hypothesis (e.g., histogram, scatterplot, boxplot, or barplot).

1d. Describe your visualization in 2-3 sentences. Does it support your hypothesis?

```{r setup, include=FALSE, echo=FALSE}
boxplot(cyst_age ~ sex)
```

### Part 3: Are monkeys with cysts more likely to die than monkeys without cysts?

1a. State a hypothesis for why monkeys with cysts are more likely to die than monkeys without cysts:

1b. State the prediction that flows from your hypothesis.

1c. Present a data visualization that tests your hypothesis (e.g., histogram, scatterplot, boxplot, or barplot).

1d. Describe your visualization in 2-3 sentences. Does it support your hypothesis?

```{r setup, include=FALSE, echo=FALSE}
tab4 = table(cyst, death)
barplot(cyst_age)
```

**IMPORTANT: A hard copy of your answers is due in class on Wednesday. For full credit, your visualizations must have clearly labeled axes and bars/boxes (i.e., it should be clear what you are showing us), and your visualizations must be relevant to the hypotheses in question. Also, remember the distinction between hypotheses and predictions!**

