---
title: "Olympic history: Women in the Olympics"
date: '2018-06-xx'
layout: post
tags: R olympics sports tidyverse
comments: true
---

```{r setup, include=FALSE}
# knitr settings
#knitr::opts_chunk$set(fig.path = "assets/Rfigs/", echo=FALSE, comment="> ", fig.align='center')
#knitr::opts_knit$set(base.dir = "/Users/nunnlab/Documents/GitHub/rgriff23.github.io/", base.url = "/")

# load packages
library("tidyverse")

# load data 
data <- read_csv("~/Documents/GitHub/Olympic_history/data/athlete_events.csv",
                           col_types = cols(
                             ID = col_character(),
                             Name = col_character(),
                             Sex = col_factor(levels = c("M","F")),
                             Age =  col_integer(),
                             Height = col_double(),
                             Weight = col_double(),
                             Team = col_character(),
                             NOC = col_character(),
                             Games = col_character(),
                             Year = col_integer(),
                             Season = col_factor(levels = c("Summer","Winter")),
                             City = col_character(),
                             Sport = col_character(),
                             Event = col_character(),
                             Medal = col_factor(levels = c("Gold","Silver","Bronze"))
                             )
                           )

# Exclude art competitions
data <- data %>% filter(Sport != "Art Competitions")

# Recode year of Winter Games after 1992 to match the next Summer Games
original <- c(1994,1998,2002,2006,2010,2014)
new <- c(1996,2000,2004,2008,2012,2016)
for (i in 1:length(original)) {
  data$Year <- gsub(original[i], new[i], data$Year)
}
```

## Introduction

The founder of the modern Olympics, Pierre de Coubertin, held a number of opinions that are at odds with the contemporary "Olympic movement". For instance, he opposed the inclusion of team sports because he insisted that the Games should be competitions among individuals. He also believed that [art](https://www.smithsonianmag.com/arts-culture/when-the-olympics-gave-out-medals-for-art-6878965/) deserved a firm place in the Games, and even won a [gold medal](https://theolympians.co/2015/12/03/pierre-de-coubertin-founder-of-the-modern-olympiad-and-gold-medalist-in-literature/) for his poem, "Ode to Sport", in the 1912 Stockholm Games. Perhaps most famously, he strongly disapproved of [women](https://www.sbs.com.au/topics/zela/article/2016/05/03/women-olympic-games-uninteresting-unaesthetic-incorrect) competing in the Olympics, calling the idea "impractical, uninteresting, unaesthetic, and incorrect". He [believed](https://www.google.com/url?sa=t&rct=j&q=&esrc=s&source=web&cd=10&cad=rja&uact=8&ved=0ahUKEwjGlr3SirvbAhUrtlkKHRt-COUQFgiIATAJ&url=http%3A%2F%2Flibrary.la84.org%2FOlympicInformationCenter%2FOlympicReview%2F2000%2FOREXXVI31%2FOREXXVI31za.pdf&usg=AOvVaw38h1u7Yfjz0ROw7XrsJ9m8) that women could participate in sports so long as they did not "make a spectacle of themselves", but that in the Olympics, "their primary role should be to crown the victors".

Coubertin was the founder and first long-term president of the IOC (1896-1925), however he was not the last IOC president to oppose the inclusion of women in the Olympics. Avery Brundage, the American president of the IOC from 1952 to 1972, is perhaps most imfamous for being a likely Nazi sympathizer, but he also expressed [doubts](https://www.independent.com/news/2013/jan/03/he-demanded-olympics-purity-not-his-own/) that women should participate in the Olympics, and argued that certain sports were simply too strenuous for women, such as the shot put or long distance runs. 

Despite resistance from many directions, women have turned up to compete in every Olympic Games aside from the first in Athens, 1896. In this post, I investigate historical trends in women's participation in the Olympics: how many there are, where they come from, and which ones find their way to the podium. 

## Data exploration

In my first [post]() analyzing the Olympic data, I showed that the number of athletes participating in the Olympics has increased dramatically over time. But how does the increase in the number of female athletes compare to the number of male athletes? 

The following interactive chart compares the growth of male and female participants over time. I exclude the non-athletic Art Competitions from the data, and I combine data from the Winter and Summer Games for each 4 year Olympiad (e.g., Sochi 2014 and Rio 2016 are grouped together). 

```{r}
# Table counting number of athletes by Year and Sex
counts <- data %>% group_by(Year, Sex) %>%
  summarize(Athletes = length(unique(ID)))
counts$Year <- as.integer(counts$Year)

# Plot number of male/female athletes vs time
plot1 <- ggplot(counts, aes(x=Year, y=Athletes, color=Sex)) +
  geom_point(size=2) +
  geom_line()  +
  scale_color_manual(values=c("darkblue","red")) +
  labs(title = "Number of male and female Olympians over time") +
  theme(plot.title = element_text(hjust = 0.5))
plot1
```

Growth in the number of female athletes largely mirrored growth in the number of male athletes up until 1996, when growth in the number of male athletes suddenly leveled off at ~8000, while the number of female athletes continued to grow at a high rate. The participation of female athletes reached its highest point during the most recent Olympiad (Sochi 2014 and Rio 2016), in which slightly more than 44%% of Olympians were women. 

But not all nations have invested equally in their female athletes: some have embraced the opportunity to win more medals in women's events, while others have been slow to include women on their Olympic teams. The following chart shows the number of female athletes versus the number of male athletes from 5 select Olympic Games (1936, 1956, 1976, 1996, and 2016), with each data point representing a National Olympic Committee (NOC) and separate best-fit regression lines for each of the 5 Games. Only NOCs represented by at least 30 athletes are included in the plot. The dashed line represents the ideal of NOCs sending teams comprised of 50% women.  

```{r}
# Count M/F/Total per country per Olympics 
# Keep only country-years with at least 30 athletes
counts_NOC <- data %>% filter(Year %in% c(1936,1956,1976,1996,2016)) %>%
  group_by(Year, NOC, Sex) %>%
  summarize(Count = length(unique(ID))) %>%
  spread(Sex, Count) %>%
  mutate(Total = sum(M,F,na.rm=T)) %>%
  filter(Total > 29)
names(counts_NOC)[3:4] <- c("Male","Female")
counts_NOC$Male[is.na(counts_NOC$Male)] <- 0
counts_NOC$Female[is.na(counts_NOC$Female)] <- 0

# Plot female vs. male athletes by NOC / Year
plot2 <- ggplot(counts_NOC, aes(x=Male, y=Female, color=Year)) +
  geom_point(alpha=0.6) +
  geom_abline(intercept=0, slope=1, linetype="dashed") +
  geom_smooth(method="lm", se=FALSE) +
  labs(title = "Female vs. Male Olympians from participating NOCs") +
  theme(plot.title = element_text(hjust = 0.5)) +
  guides(color=guide_legend(reverse=TRUE))
plot2
```

The chart shows that although there wasn't much change from 1936 to 1956, there was dramatic improvement in female participation from 1956 to 2016. In 1996 and 2016, some NOCs even sent a majority of female athletes to the Games (these are represented by points above the dashed line). 

So which NOCs are leading the way for gender equality in the Olympics? The following charts rank nations by the ratio of female to male athletes on their Olympic Teams. Like the previous chart, an NOC must have sent at least 30 athletes to the Games to be considered in the chart. 

```{r}

```

[analysis]

The final question I ask here is: which nations won the most medals in women's events? 

```{r}

```

[analysis]


