---
title: "Olympic history (part 2): trends in international participation and medalling"
date: '2018-xx-xx'
layout: post
tags: R olympics sports tidyr ggplot2
comment: yes
---

```{r setup, include=FALSE}
# knitr settings
#knitr::opts_chunk$set(fig.path = "assets/Rfigs/", message=FALSE, comment="> ", fig.align='center')
#knitr::opts_knit$set(base.dir = "/Users/nunnlab/Desktop/GitHub/rgriff23.github.io/", base.url = "/")
```

This is the second in a series of posts analyzing data on the modern Olympic Games, from Athens 1986 to Sochi 2014. Here, I show how international participation in the Games has grown over time and identify which nations have dominated the podium throughout history. As in the other posts in this series, I use the `tidyverse` R package for manipulating and plotting data. 

To start, load the packages `ggplot2` and `tidyverse`, and the dataset `athlete_events` (which can be found in the [GitHub repo](https://github.com/rgriff23/Olympic_history). 

```{r, message=FALSE}
# load packages
library("tidyverse")

# load data 
athlete_events <- read_csv("~/Desktop/GitHub/Olympic_history/athlete_events.csv",
                           col_types = cols(
                             Name = col_character(),
                             Sex = col_factor(levels = c("M","F")),
                             Birthplace = col_character(),
                             Age =  col_integer(),
                             City = col_character(),
                             Sport = col_character(),
                             NOC = col_character(),
                             Medal = col_factor(levels = c("Bronze","Silver","Gold")),
                             Year = col_integer(),
                             Season = col_factor(levels = c("Summer","Winter"))
                           ))
```

The first modern Olympics, Athens 1986, included 216 athletes from 50 nations competing in 34 summer events. Since then, the Olympics has grown to include 10,673 athletes from 214 nations competing in 56 events in the 2012 London Olympics, and 2,883 athletes from 162 nations competing in 50 events in the 2010 Vancouver Olympics. The following charts depict the number of athletes, nations, and events in the Summer and Winter Games over time:

```{r num_athletes_vs_time}
# plot number of athletes vs time
athletes_time <- athlete_events %>% select(Name, Year, Season) %>% group_by(Year,Season) %>% summarize(Athletes = length(unique(Name)))
ggplot(athletes_time) + geom_line(aes(x=Year, y=Athletes, color=Season))
```
 
```{r num_nations_vs_time}
# plot number of nations vs time
nations_time <- athlete_events %>% select(NOC, Year, Season) %>% group_by(Year,Season) %>% summarize(NOCs = length(unique(NOC)))
ggplot(nations_time) + geom_line(aes(x=Year, y=NOCs, color=Season))
```

```{r num_sports_vs_time}
# plot number of sports vs time
sports_time <- athlete_events %>% select(Sport, Year, Season) %>% group_by(Year,Season) %>% summarize(Sports = length(unique(Sport)))
ggplot(sports_time) + geom_line(aes(x=Year, y=Sports, color=Season))
```

Now let's look at which nations have historically dominated the medal count in the Olympics. I don't want to distinguish between gold, silver, and bronze, so I will add a column to the data called `Medal2` which contains a logical (TRUE/FALSE) indicator of whether a medal was won:

```{r}
# add logical column for medaling
athlete_events <- mutate(athlete_events, Medal2 = !is.na(Medal))
```

Typing `sum(athlete_events$Medal2)` reveals 37,774 medals in the dataset. 

I also want to combine certain National Olympic Committee (NOC) codes because they represent roughly the same geographic area and population. Specifically, I will reclassify East and West Germany as simply "GER", and the Soviet Union and [Unified Team](https://en.wikipedia.org/wiki/Unified_Team_at_the_Olympics) as "RUS".

```{r}
# combine NOCs for Germany and Russia
athlete_events$NOC[athlete_events$NOC%in%c("FRG","GDR")] <- "GER"
athlete_events$NOC[athlete_events$NOC%in%c("URS","EUN")] <- "GER"
```

Now let's take a peek at how nations stack up in terms of historical medal counts by creating a new nation-level dataframe:

```{r}
# create tibble of historical medal counts for each nation
medal_count <- athlete_events %>% group_by(NOC) %>% summarize(n = sum(Medal2)) %>% arrange(desc(n)) %>% mutate(proportion = round(n/sum(n),3))

# view all nations with at least 1% of the medal count
medal_count %>% filter(proportion >= 0.01) %>% print(n=Inf)

# compute proportion of medals won by the top 10 countries
sum(medal_count$proportion[1:10])

# compute proportion of medals won by nations with at least 1% of the medal count
filter(medal_count, proportion >= 0.01) %>% select(proportion) %>% sum
```

Germany comes out on top with a whopping 16.9% of all medals in Olympics history. The top 10 nations have won a combined 61.2% of all medals. Only 27 nations have at least 1% of the total, and these 27 nations have won a combined total of 89.6% of all medals in history. 

This birds eye view might obscure interesting temporal patterns and differences between the Summer and Winter Games. For a more nuanced view, I use line graphs to reveal historical trends in the proportion of medals won in past Olympics, plotting Summer and Winter Games data separately. 

Before plotting, I have to do some data manipulation. First, I must produce separate dataframes for the Summer and Winter data. Second, I need to compute the proportion of medals each country won in each Games. Third, to avoid overly cluttered plots, I want to subset each dataframe to include only data for the top 5 medal counts in each Olympics. Here is the code for the Summer Games data:

```{r}
# filter to summer data and select only necessary columns
summer <- athlete_events %>% filter(Season == "Summer") %>% select(NOC, Year, Medal2)

# count medals for each NOC-year, total medals for each year, and proportion for each NOC-year
summer_noc_counts <- summer %>% group_by(Year, NOC) %>% summarize(medals = sum(Medal2))
summer_totals <- summer %>% group_by(Year) %>% summarize(total_medals = sum(Medal2))
summer_props <- summer_noc_counts %>% left_join(summer_totals, by="Year") %>% mutate(proportion = round(medals/total_medals,3))

# filter to top 10 countries from each year
summer_top10 <- summer_props %>% group_by(Year) %>% filter(rank(desc(proportion), ties.method="random") <= 5) %>% select(Year, proportion, NOC)

# view all unique nations in dataset
summer_top10$NOC %>% unique # 19 countries
```

Here is the same code for the Winter Games data, without all the comments: 

```{r}
# preoare winter data
winter <- athlete_events %>% filter(Season == "Winter")
winter_noc_counts <- winter %>% group_by(Year, NOC) %>% summarize(medals = sum(Medal2))
winter_totals <- winter %>% group_by(Year) %>% summarize(total_medals = sum(Medal2))
winter_props <- winter_noc_counts %>% left_join(winter_totals, by="Year") %>% mutate(proportion = round(medals/total_medals,3))
winter_top10 <- winter_props %>% group_by(Year) %>% filter(rank(desc(proportion), ties.method="random") <= 10) %>% select(Year, proportion, NOC)
winter_top10$NOC %>% unique # 22 countries
```

There are 19 countries in the Summer Games data and 22 countries in the Winter Games data.

There are so many countries that it is difficult to distinguish them all using different line colors, so I want to add country flags as plotting symbols using the `emojifont` package. For this to work, I need to add a data column that specifies the emojis to use to plotting symbols. 

```{r}
# load emojifont
library("emojifont")

# create table matching each NOC with a flag
flags <- tibble(
  NOC = c(summer_top10$NOC, winter_top10$NOC) %>% unique,
  flag = c("U+1F1E6 U+1F1FA",
           "U+1F1E8 U+1F1F3",
           "U+1F1E9 U+1F1F0",
           "U+1F1EB U+1F1F7",
           "U+1F1EC U+1F1E7",
           "U+1F1E9 U+1F1EA",
           "U+1F1EC U+1F1F7",
           "U+1F1ED U+1F1FA",
           "U+1F1F3 U+1F1F1",
           "U+1F1FA U+1F1F8",
           "U+1F1E7 U+1F1FE",
           "U+1F1E8 U+1F1E6",
           "U+1F1EE U+1F1F9",
           "U+1F1E8 U+1F1ED",
           "U+1F1E8 U+1F1FA",
           "U+1F1EB U+1F1EE",
           "U+1F1F3 U+1F1F4",
           "U+1F1F7 U+1F1F4",
           "U+1F1F8 U+1F1EA",
           "U+1F1EE U+1F1F3",
           "U+1F1EF U+1F1F5",
           "U+1F1E6 U+1F1F9",
           "U+1F1E8 U+1F1FF",
           "U+1F1F5 U+1F1F1",
           "U+1F1E7 U+1F1EC",
           "U+1F1F0 U+1F1F7",
           "U+1F1F7 U+1F1FA",
           "U+1F1E7 U+1F1F7"
           )
  )

# add flag to winter and summer data
summer_top10 <- summer_top_10 %>% left_join(flags, by="NOC")
winter_top10 <- winter_top_10 %>% left_join(flags, by="NOC")
```

Now we can plot the data, starting with the Summer Games:

```{r}
# plot summer data
ggplot(summer_top10) +
  geom_line(aes(x=Year, y=proportion, group=NOC, color=NOC))
```

[comments]

Now let's plot the Winter Games data: 

```{r}
# plot winter data
ggplot(winter_top10) +
  geom_line(aes(x=Year, y=proportion, group=NOC, color=NOC))
```

[comments]

[concluding remarks]

Up next: Historical trends in women's participation in the Olympics.

**Footnotes**

For a comprehensive history of the modern Olympics, I recommend ['Politics, Power, and the Olympic Games'](https://www.amazon.com/Power-Politics-Olympic-Games-Alfred/dp/0880119586/ref=sr_1_1?ie=UTF8&qid=1523105541&sr=8-1&keywords=politics+power+olympic+games) by Alfred Senn. It's a bit dry, but very thorough.

