---
title: "Spatial co-distribution of sasquatch and UFO sitings in the U.S."
date: '2018-04-xx'
layout: post
tags: R GIS tutorial
comment: yes
---

```{r setup, include=FALSE}
# knitr settings
#knitr::opts_chunk$set(fig.path = "assets/Rfigs/", echo=FALSE, comment="> ", fig.align='center')
#knitr::opts_knit$set(base.dir = "/Users/nunnlab/Desktop/GitHub/rgriff23.github.io/", base.url = "/")

# load packages 
library("ggplot2")
```

For some time now, I've been looking forward to upgrading my R GIS skills to include `ggplot2` and `ggmap` graphics. In this post, I use `ggplot2` and `ggmap` to investigate the spatial occurrence and co-occurrence of historical sasquatch and UFO sitings in the U.S. 

I've actually had this data laying around since 2015, but forgot about it until recently when I was cleaning up files on my computer. The sasquatch data was downloaded from the [Bigfoot Field Researchers Organization](http://www.bfro.net/) website. The UFO data, which was originally compiled from the [National UFO Reporting Center](http://www.nuforc.org/), was downloaded from [GeoCommons](http://geocommons.com/). This particular UFO dataset has since been removed from GeoCommons, but it is similar to several versions that can be found floating around on the internet at the time of this post, including on [data.world](https://data.world/timothyrenner/ufo-sightings) and [kaggle](https://www.kaggle.com/camnugent/ufo-sightings-around-the-world). 

### Preparations

```{r}
# load packages
library(maptools)
```

The original files are in kml format, so you can extract long-lat coordinate data using the function `getKMLcoordinates` from the `maptools` package: 

```{r}
# extract long-lat coordinates from kml files
sasquatch <- getKMLcoordinates("sasquatches.kml", ignoreAltitude=TRUE)
ufo <- getKMLcoordinates("ufo.Kml", ignoreAltitude=TRUE)
```

This takes quite a while for the UFO data - about 45 minutes on my machine. To speed this up in the future, I created matrices with the long-lat coordinates and exported them as csv files:

```{r}
# convert list to matrix of long-lat coordinates
sasquatch <- matrix(unlist(sasquatch), ncol=2, byrow=TRUE) # 7770 sitings
ufo <- matrix(unlist(ufo), ncol=2, byrow=TRUE) # 61088 sitings
colnames(ufo) <- colnames(sasquatch) <- c("long","lat")

# export long-lat matrices
write.csv(sasquatch, "sasquatch.csv", row.names=FALSE)
write.csv(ufo, "ufo.csv", row.names=FALSE)
```

These csv files are available in the [GitHub repo]() for this analysis, so you can skip the steps above and import the coordinate matrices from the repo directly into R:

```{r}
### UPLOAD DATA AND FIX THESE LINKS
# import csv files with long-lat coordinates
sasquatch <- read.csv("sasquatch.csv")
ufo <- read.csv("ufo.csv")
```

Checking the size of these matrices with `nrow(sasquatch)` and `nrow(ufo)` reveal 7770 sasquatch sitings and 61088 UFO sitings, making UFO sitings nearly 8 times more common than sasquatch sitings. 

### Plotting the (co)distribution of sasquatch and UFO sitings

First, I'll plot the locations of sasquatch and UFO sitings separately on a map of the U.S. Since multiple sitings often occur at the same coordinates, I will make the size of the points proportional to the number of sitings at each location. First, for sasquatches:

```{r}
# sasquatches
```

[summary] Now for UFOs.

```{r}
# UFOs
```

[summary][describe why/how to plot co-distribution]

```{r}
# co-distribution of sasquatches and UFOs
```

[summary]
