---
layout: post
title: "btw update: BayesTraitsV3 and Windows compatibility"
date: 2017-06-xx
tags: R BayesTraits btw 
comment: true
---

I finally updated <a target="_blank" href="https://github.com/rgriff23/btw">`btw`</a> to run <a target="_blank" href="https://www.evolution.rdg.ac.uk/BayesTraits.html">BayesTraitsV3</a> from R on both Mac OS and Windows systems. Thank you to <a target="_blank" href="http://dk-giscience.zgis.net/index.php/37-website/256-ranacher-peter">Peter Ranacher</a> for contributing code to run `btw` functions on Windows. 

The updated `btw` supports all the new functionality of BayesTraitsV3, including tagging nodes and modeling distributions of tip data. Any commands you can pass to BayesTraitsV3, you can pass to the `btw` functions. 

In this iteration of `btw`, I decided to make user input more flexible. There are fewer arguments, and users can now supply a vector containing any BayesTraits commands they wish. These commands are copied as-is into the `input.txt` file. In addition to adding flexibility for users, I think this format makes it easier for anyone who is familiar with BayesTraits to learn to use `btw`.

I also endeavored to be thorough in including output files in the `btw` results. In the previous version of `btw`, I was lazy and left a few things out (e.g., Schedule files), and I got a couple e-mails about it. I tried not to make the same mistake this time around.

The updated [tutorial](/projects/btw) on my projects page shows how to access the functionality of BayesTraitsV3 with `btw`. Install the new version just like to old version:

```{r, eval=FALSE}
library("devtools")
install_github("rgriff23/btw")
```

And if you ever want to install the older version of `btw` to work with BayesTraitsV2, you can do so like this:

```{r, eval=FALSE}
install_github("rgriff23/btw", ref="v1")
```


